---
title: "R Small Group: Class 1"
author: "Amy Allen & Dayne Filer"
date: "June 10, 2016"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

## Introduction to R
The most basic introduction to R is through simple math calculations. For example:

```{r}
3 + 4
```

Similarly, R can do subtraction, multiplication, and division:

```{r}
4 - 2
2*120
1e3/2
```

All these calculations are taking place in your global R environment. (Although it is beyond what you need to understand for this class, you can check which environment you are working in by running `environment()`; the result should be: `<environment: R_GlobalEnv>`.) The global R environment can store variables defined by the user.

The `<-` command is used to store variables in R. For example, if we want to store the result of `3 + 4` to use in future calculations, we could do something like the following:

```{r}
x <- 3 + 4
```

Notice, code did not return any result. The result is now stored in `x`. In the example above, "x" is the name of the variable. The `<-` operator told R to calculate `3 + 4` and store the results in a variable called `x`. You can now access the variable by typing it into the console:

```{r}
x
```

The `x` variable can also be used in calculations and to create new variables.

```{r}
y <- 2 + x
y
```

It can be hard to remember which variables are available. R provides the `ls` function for displaying the variables contained in your environment. Functions have names, in this case `ln`. What happens when you run `ls`? 

```{r}
ls
```

The output shows the code behind the `ls` function. To use a function you have to add `()`.

```{r}
ls()
```

Running `ls()` returns `[1] "x" "y"`, indicating there are two variables, `x` and `y`, in the global environment. Here we see a new type of result: the `x` and `y` are surrounded by quotation marks. The quotation marks indicate a different type of data. Programming languages include different data classes. While R includes many different types, we will discuss the three most common: (1) numeric, (2) character, and (3) logical. We can check the data class with the `class` function. 

```{r}
vars <- ls()
class(x = vars)
class(x = x)
```

Notce the `class` function requires an input, in this case called "x". Running `class()` without any input will give an error message:

```{r, error=TRUE}
class()
```

In some programming languages the user has to specify the data class beforehand. R makes an educated guess on the data class, which is convenient, but can cause unexpected problems if the user is not careful. Let's expolore data classes further after making some new variables. 

```{r}
var1 <- c(1, 2, 3)
var2 <- c("1", "2", "3")
var3 <- c("a", "b", "c")
```

The `c` function creates a vector -- or a string of values. We did not tell R what data class to make the vectors. Run `class` to see how R interpreted the vectors. 

```{r}
class(var1)
class(var2)
class(var3)
```

Even though `var2` could be interpreted as numeric, adding the quotation marks created a character vector. When combining or comparing two values/vectors R attempts to coerce one to the higher order data class. From the R documentation:

> If the two arguments are atomic vectors of different types, one is coerced to the type of the other, the (decreasing) order of precedence being character, complex, numeric, integer, logical and raw.

Although we won't discuss all of the data classes listed above, it provides a nice refernce for future use. (To find help in the documentation play around with `?` and `??`. `?` will show the documentation for a specific function, e.g. `?mean`, and `??` searches the documentation for a topic, e.g. `??average`. You can see from the `??average` search that a quick Google search will often provide better results when you don't know the name of the function you are looking for.)



R is composed of functions and variables, i.e. everything in R is either a function or a variable. Functions and variables all have names, for example: `ls`. 

Execute your first command! R provides the `ls` function for displaying the variables contained in your environment. 


