---
title: "R Small Group: Class 1"
author: "Amy Allen & Dayne Filer"
date: "June 10, 2016"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to R
The most basic introduction to R is through simple math calculations. For example:

```{r}
3 + 4
```

Similarly, R can do subtraction, multiplication, and division:

```{r}
4 - 2
2*120
1e3/2
```

All these calculations are taking place in your global R environment. (Although it is beyond what you need to understand for this class, you can check which environment you are working in by running `environment()`; the result should be: `<environment: R_GlobalEnv>`.) The global R environment can store variables defined by the user.

The `<-` command is used to store variables in R. For example, if we want to store the result of `3 + 4` to use in future calculations, we could do something like the following:

```{r}
x <- 3 + 4
```

Notice, code did not return any result. The result is now stored in `x`. In the example above, "x" is the name of the variable. The `<-` operator told R to calculate `3 + 4` and store the results in a variable called `x`. You can now access the variable by typing it into the console:

```{r}
x
```

The `x` variable can also be used in calculations and to create new variables.

```{r}
y <- 2 + x
y
```

It can be hard to remember which variables are available. R provides the `ls` function for displaying the variables contained in your environment. Functions have names, in this case `ln`. What happens when you run `ls`? 

```{r}
ls
```


R is composed of functions and variables, i.e. everything in R is either a function or a variable. Functions and variables all have names, for example: `ls`. 

Execute your first command! R provides the `ls` function for displaying the variables contained in your environment. 


