---
title: 'Homework 3: Solutions'
author: "Amy Allen & Dayne Filer"
date: "June 28, 2016"
output:
  pdf_document:
    highlight: tango
  html_document: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

<!-- Here we style out button a little bit -->
<style>
  .showopt {
    background-color: #004c93;
    color: #FFFFFF; 
    width: 100px;
    height: 20px;
    text-align: center;
    vertical-align: middle !important;
    border-radius: 8px;
    float:right;
    }

    .showopt:hover {
    background-color: #dfe4f2;
    color: #004c93;
  }

</style>

<!--Include script for hiding output chunks-->
<script src="hideOutput.js"></script>

Load data:

```{r}
setwd("~/Documents/rclass")
data <- read.csv("heights.csv", header = TRUE)
```

1. Make a histogram of height using the `hist()` function. Include an appropriate title and axis labels.

```{r}
hist(data$Height, 
     xlab = "Height (inches)", 
     ylab = "Frequency", 
     main = "Histogram of Height")
```

\pagebreak
2. Now make the same histogram but include 20 bins and make it your favorite color. 

```{r}
hist(data$Height, xlab = "Height (inches)", 
     ylab = "Frequency", 
     main = "Histogram of Height", 
     breaks = 20, 
     col = "purple")
```

\pagebreak
3. Make a bar chart of how many people are left or right handed using the `barplot()` function. Hint: `data$Handedness[data$Handedness=="Left"]` will return a vector including on those entries where `data$Handedness` is `"Left"` and `length()` will return the length of a vector. 

```{r}
barplot(height = c(length(data$Handedness[data$Handedness =="Left"]),
                   length(data$Handedness[data$Handedness =="Right"])), 
        names.arg = c("Left Handed","Right Handed"), 
        ylab = "Count", 
        main = "Handedness")
```

\pagebreak
4. Make a bar chart of how many people have the following eye colors: blue, brown, other. Make the color of each bar correspond to the eye color (you can choose any color for other). 

```{r}
barplot(height = c(length(data$Eye.Color[data$Eye.Color == "Blue"]),
                   length(data$Eye.Color[data$Eye.Color == "Brown"]),
                   length(data$Eye.Color[data$Eye.Color == "Other"])), 
        names.arg = c("Blue","Brown","Other"), 
        ylab = "Count", 
        main = "Eye Color",
        col = c("blue","brown","purple"))
```

\pagebreak
5. Make a box plot of the height using the `boxplot()` function. 

```{r}
boxplot(x = data$Height, 
        ylab = "Height (inches)", 
        main = "Boxplot of Height")
```