---
title: "Reading & Cleaning Data"
output: 
  html_notebook
---
 
# Topics:
 - discuss `readr::read.csv` and why it's better than `read.csv`
 - `readxl::read_xls` to show how to import excel tables
 - introduce `skip` and `comment` parameters
 - discuss regular expression **basics** & `gsub` command for replacing characters

```{r setup}
library(magrittr)
library(dplyr)
```

# Download example dataset:
```{r}
link <- "https://ucr.fbi.gov/crime-in-the-u.s/2015/crime-in-the-u.s.-2015/tables/table-9/table_9_offenses_known_to_law_enforcement_by_state_by_university_and_college_2015.xls"
file <- "crime.xls"
download.file(link, file)
```

```{r}
read_xls(file)
```
Looks like the data we want is actually a litter lower down in the file. 
We can skip over lines at the top with the `skip` parameter.
```{r}
(crime <- read_xls(file, skip = 3))
```
Even though it's good to look at the **top** of the data, you should also look at the **bottom**! This is often overlooked and has caused problems for many people in the past. You can do this easily with `tail()`.
```{r}
tail(crime)
```

Lo' and behold, this doesn't look right.

Looks like the last 8 rows of the data.frame are not part of our data. We should remove these.
```{r}
tail(crime, 20)
```
I like using base subsetting for these problems since usually the reason you're
removing rows at the bottom is because a tool will output extra information at
the bottom of an output file. In case you rerun the tool later with more data
and the length of the file changes, you can't rely on reading in a set number of
lines. If the tool always outputs the same number of lines at the end of the
file, it's safer to just drop the last `n` lines from the data after reading.
We'll do this here by removing the last 8 lines.
```{r}
keep <- 1:(nrow(crime) - 8) # create a vector of rows to keep
crime <- crime[keep,] # subset by row
```

Now our data is correct at the top and the bottom.
```{r}
head(crime)
```
```{r}
tail(crime)
```

Let's take another close look at the dataset:
```{r}
str(crime)
```

Note that the `X__1` through `X__4` columns look empty. We can check that by looking at the unique values in each column:
```{r}
crime$X__1 %>% unique
crime$X__2 %>% unique
crime$X__3 %>% unique
crime$X__4 %>% unique
```
Retyping is annoying. We can use pattern matching to select these all at once, and looking at their `summary`.

We can see that they only contian `NA` values, so there's no reason to keep them.
```{r}
crime %>% 
  select(matches("X_")) %>% 
  summary()
```

We can drop them by matching a pattern as above using `-`.
```{r}
crime %<>% 
  select(-matches("X_")) # drop any columns that match the pattern "X_"
```

Check that the columns were dropped. Also note that the remaining column names
are poorly formatted.

Whats the deal with the `\n`?
```{r}
names(crime)
```

`\n` along with `\t` and `\r` are special characters that often appear in files.
`\n` and `\r` are codes that stand in place for a newline (like pressing enter).
`\r` is called the 'carriage return' character which is usually only added when
files are created using a Windows OS. `\t` stands for a tab.

In the above example the column names have newline characters in them, which
will make it impossible to reference by name since newlines are sometimes
meaningful in R.

For example:
Try to autocomplete the following:
```{r}
# Rstudio should autocomplete this:
crime$`Murder and
nonnegligent
manslaughter`
# But it will return an error because of the newlines (it will run if you highlight the whole thing and run the code)
```
```{r}
# this won't work either because the column name requires newlines
crime$`Murder and nonnegligent manslaughter`
```

We need a way to replace newline characters found in all colnames of `crimes`. 
One easy way to do this is with **regular expressions**.

**Regular expressions** (also called **Regex's**) are a way of matching patterns in text.
The `gsub` command will use regular expressions to match a pattern and replace it for a different one.

Here are a few simple examples:
```{r}
gsub("Bob", "Tom", "Bob and Sally") # Replace Tom for Bob in the string "Bob and Sally"
```

```{r}
myString <- "My Birthday is January 01"
gsub("01", "10", myString)
```

```{r}
gsub(" ", "_", myString) # replace all spaces for _
```

## rename with regex
**Note to teachers:** This is a good opportunity to let the students play with
the code on their own to get these solutions.

Lets first drop out the `\n`
```{r}
names(crime) %>%
  gsub("\n", "_", .)
```
When renaming colnames in batch like this, I like to string together all my
regexes and monitor the output as I add them on to make sure things make sense,
then I'll go back and save the changes once I'm happy with the results. We'll
take that approach here by continually adding new `gsub` commands until we get
colnames we like.

Now that we've solved the newline issue, what other issues exist? Take a look at
the `/`. This is a bad character in a colname because R needs to escape it to
access it. Let's change it to '.'

```{r}
names(crime) %>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) 
```

While we're at it, let's also drop out the `-` because this means "subtract" in
R, so it is also a bad character for colnames. We'll just delete it.
To delete things with regex's, you can replace a pattern with an empty string: ""
```{r}
names(crime) %>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) %>% 
  # replacing for an empty string will delete a character/pattern
  gsub("-", "", .) 
```

There are also some spaces in the colnames (look carefully!). We don't want those either, so change them to "_".
```{r}
names(crime) %>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) %>% 
  gsub("-", "", .) %>% 
  gsub(" ", "_", .)
```

You'll also notice that some columns end in numbers. These aren't descriptive, so let's delete them too. 

We can use the special regex character `\d` which means "match any number".
Because R is weird, you have to escape the first `\` when you call `\d` in
regex's. This is often true for other regex characters like '.' (match any
character) which has to be written as `\\.` in R. This might not make sense
right now, but if you work more with regex's later, keep this in mind!

```{r}
names(crime) %>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) %>% 
  gsub("-", "", .) %>% 
  gsub(" ", "_", .) %>% 
  gsub("\\d", "", .)
```

Next, let's get rid of the parentheses. We can use another regex pattern `[]`
which will match any characters inside the square brackets to delete both "(" and ")".
```{r}
names(crime) %>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) %>% 
  gsub("-", "", .) %>% 
  gsub(" ", "_", .) %>% 
  gsub("\\d", "", .) %>% 
  gsub("[()]", "", .)
```

Finally, let's get rid of the capitalization in the colnames to make it
consistent. We'll convert everything to lowercase with the function `tolower()`.
```{r}
names(crime) %>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) %>% 
  gsub("-", "", .) %>% 
  gsub(" ", "_", .) %>% 
  gsub("\\d", "", .) %>% 
  gsub("[()]", "", .) %>% 
  tolower()
```
Now save the new names with `%<>%` which will send the final output back as input.
```{r}
names(crime) %<>%
  gsub("\n", "_", .) %>% 
  gsub("/", ".", .) %>% 
  gsub("-", "", .) %>% 
  gsub(" ", "_", .) %>% 
  gsub("\\d", "", .) %>% 
  gsub("[()]", "", .) %>% 
  tolower()
```

```{r}
names(crime)
```

For more reading on regex's check:
[Regex cheatsheet](http://regexlib.com/(X(1)A(xSygfbbpgXfgBpYdzZKZbaM80NvGWFKqz8H3p14z0goMZmVZU6P1V8S_EcyNeeY0ryvSsa9Ndw9_54dwIXQI5fjTdwJzOo1NhOnVqmLTuZ1NDcdA-UseLK7W01hB-KYIvuVOGgWiImZ-T7mf0uWFBNQbtY1KXM9TgHosz31NGuevvup3lyGMYJLwdgybbcAg0))/CheatSheet.aspx)

# Fill in missing values
Take a look at the `state` and `university.college` columns. Each row labeled as
`NA` takes the value of the previously assigned row. We need a way to fill these values down.
```{r}
crime
```

To do this, we can use the `tidyr` package, whose function `fill()` will do just this:
```{r}
library(tidyr)
(crime %<>% 
  fill(state, university.college))
```

## Make a plot with your new tidy data!
```{r}
library(ggplot2)
crime %>% 
  ggplot(aes(student_enrollment, violent_crime + property_crime)) +
    geom_point(aes(color = state)) + 
    theme(legend.position = "none") +
    ylab("Crimes per campus")
```

## Save the cleaned data:
It's always a good idea to save the clean data for easy sharing with others later.
```{r, eval = F}
readr::write_csv(crime, "crime_cleaned.csv")
```

```{r cleanup, include=F, echo=F, message=F, warning=F}
file.remove(file)
```

# Additional Import Functions
The `readr` package [documentation here](https://readr.tidyverse.org/index.html)
provides several user-friendly functions for reading comma-separated (`.csv`
with `readr::read_csv()`), tab-separated (`.tsv` with `readr::read_tsv()`), and
other delimited files (`readr::read_delim()`). You should prefer these functions
over base-R implementations like `read.csv()` or `read.delim()`, as the `readr`
functions have more consistent and reproducible behavior across filesystems, and
they use more sensible defaults.

