---
title: 'R Small Group: Class 2'
author: "Amy Allen & Dayne Filer"
date: "June 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```


1. Create a data frame with two columns: "col1" with the letters a to f, "col2" with the numbers 1 to 6, and "col3" with the alternating TRUE and FALSE values. Store the data frame as "mydf."

```{r}
mydf <- data.frame(col1 = c("a", "b", "c", "d", "e", "f"), 
                   col2 = 1:6, 
                   col3 = c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE))
mydf
```

2. Coerce `mydf` from exercise 1 to a matrix using the `as.matrix` function. Predict how the data will change. What class will the data be? Will there be column names? Row names?
```{r}
as.matrix(mydf)
```

3. Subsetting can also be done with logical vectors the same length of the object or dimension you would like to subset. For example, if you have a list `l` with three elements, `l[c(TRUE, FALSE, TRUE)]` would return a list with elements one and three from `l`. Figure out how to use `mydf` to subset only to rows where 'col3' is `TRUE`. (You should attempt to think of a solution that only requires one line of code. Hint: you can pass elements of an obect to itself.)
```{r}
mydf[mydf$col3, 1:3]
```

\pagebreak

4. You can stack subsetting operators next to each other. Using `l2` from above, select the 7 from 'vec.' Now try to select the last two rows from 'mat'. (Again, you should attempt to think of a solution that only requires one line of code for each selection.)
```{r}
l2 <- list(vec = c(1, 3, 5, 7, 9),
           mat = matrix(data = c(1, 2, 3), nrow = 3))
l2
# 7 from 'vec'
l2$vec[4]

# last two rows from 'mat'
l2$mat[2:3, ]
```

5. Recall how you added an element to `l2` using the `$` operator. Again, a data frame is just a special list. Add a column to `mydf` called "col4" with a vector of your choice.
```{r}
mydf$col4 <- c(1, 4, 9, 16, 25, 36)
```